Using Overlapping Communities and Network Structure for Identifying Reduced Groups of Stress Responsive Genes

### 2020-12-07 camila/report/case.tex
bmc version

\section{Case Study}
\label{sec.case}
This section presents a case study on the identification of genes that may respond to saline stress according to the methodology presented in section~\ref{sec. framework}. 
The RNA-seq data was obtained from GEO database
\cite{GEOAcces90:online} (accession number GSE98455). This data
corresponds to $n_0=57845$ gene expression profiles of shoot tissues
measured for both control and salt condition in $m=92$ accessions of
the Rice Diversity Panel 1, with $r=2$ biological replicates. A total
of $p=3 $ phenotypic traits are used: shoot $K^+$ content, root
biomass and shoot biomass. These traits were measured for the same
$92$ genotypes, under control and salt stress conditions, and can be
found in the supplementary information
of~\cite{campbell2017allelic}.

\subsection{Data pre-processing}

DESeq2 normalization is applied to the raw data and the biological
replicates are averaged. Genes exhibiting low variance are
identified as those with ratio of upper quantile to lower quantile
smaller than $1.5$ and are removed from the normalized data. Genes
with low expression, corresponding to those having more than $80\%$
samples with values smaller than $10$, are also removed. A total of
$n_1 = 8928$ genes are kept after this filtering process.

From the Log Fold Change matrix $L_0$, genes whose difference between
upper quantile and lower quantile greater than $0.25$ are
removed. Therefore, the resulting matrix $L_1$ contains the log ratios
of $n_2 = 8928$ genes. The logarithmic ratios of the phenotipic data,
for the $92$ accessions and the $3$ traits, is also computed.

\subsection{Co-expression network construction}

The Log Fold Change matrix $L_1$ is used to compute the corresponding
similarity matrix.  For this network, it is observed that $\beta=3$
is the smallest integer such that the $R^2 \geq 0.8
$. Figure~\ref{fig:beta} depicts the degree distribution of the
similarity matrix (left) and the degree distribution of the adjacency
matrix (right), which is the degree distribution of a scale-free
network with $R^2 = 0.8$ with $\beta = 3$.

%scale-free beta 1 y 3
\begin{figure}[htpb]
  \centering
    \includegraphics[clip,width=0.8\textwidth]{figures/pick_beta.png}
  \caption{Degree distribution of $A$ (left) and $\hat{A}$ (right)}
  \label{fig:beta}
\end{figure}

The resulting adjacency matrix $A$ represents a complete graph
$G=(V,E)$, with $|V| = 8928$ genes and $|E| = 39850128$ edges.

\subsection{Co-expression module identification}

The adjacency matrix $A$ is transformed into an unweighted network 
$\hat{A}$ applying the approach described in~\cite{aoki2007approaches}, 
based on the density of the network combined with the decreasing number
of nodes and edges with higher PCC values. The cutoff value is set to $0.2$,
thus keeping only the connections above this threshold and removing the 
isolated nodes. The resulting adjacency matrix $\hat{A}$ has $5810$ connected 
genes and accounts for $16875145$ edges.

After applying the HLC algorithm, a total of $4131$ genes are
distributed in $c = 5143$ overlapping modules of at least $3$ genes. Figure~\ref{fig:overlap} presents a histogram of the overlapping
percentage of these genes, measured as the proportion of modules to
which each gene belongs. The first bar of the histogram represents the
genes with zero overlap, corresponding to $28\%$ of the total
genes; remaining $72\%$ represents the genes belonging to more than one
module.

%overlapping percentage
\begin{figure}[htpb]
  \centering
    \includegraphics[clip,width=0.5\textwidth]{figures/artificial_modules.png}
  \caption{Overlapping percentage}
  \label{fig:overlap}
\end{figure}

\subsection{Module association to phenotypic traits}

The phenotypic traits under study are shoot $K^+$ content, root
biomass and shoot biomass. Figure \ref{fig:pdata} suggests that there
are significant differences in the values of these phenotypic traits
between stress and control conditions. This supports the working
hypothesis that these three variables represent tolerance-associated
traits in rice under salt stress.

% boxplots
\begin{figure}[htpb]
  \centering
    \includegraphics[clip,width=1\textwidth]{figures/phenotypic_traits.png}
  \caption{Phenotipic traits distribution under control and salt stress.}
  \label{fig:pdata}
\end{figure}

Using the affiliation matrix $F$ derived from the HLC output and the
Log Fold Change matrix $L_1$, a matrix $M$ is built by computing the
egengene for each of the $c = 5143$ modules. The LASSO technique is
applied by using each of the phenotypic traits as the outcome
variable, one at a time. As shown in Figure~\ref{fig:cross-val},
cross-validation is performed for each phenotypical trait in order to
select the corresponding regularization parameter $\lambda$ that
minimizes the mean-squared error.

\begin{figure}[htpb]
  \centering
  \subfloat[$K$ shoot, $\lambda = 0.013$]{%
    \includegraphics[width=0.33\textwidth]{figures/lambda_kshoot_2.png}
    \label{fig:lambda_kshoot}
  }
  \subfloat[Shoot biomass, $\lambda = 0.025$]{%
    \includegraphics[width=0.33\textwidth]{figures/lambda_BMshoot_2.png}
    \label{fig:lambda_BMshoot}
  }
  \subfloat[Root biomass, $\lambda = 0.035$]{%
    \includegraphics[width=0.33\textwidth]{figures/lambda_BMroot.png}
    \label{fig:lambda_BMroot}
  }
  \caption{Cross-validation of the LASSO regularization parameter
    $\lambda$, for each phenotypic trait.}
  \label{fig:cross-val}
\end{figure}

Finally, three LASSO models are adjusted by using the corresponding
$\lambda$ and phenotypical data with the egengens of matrix $M$. As
result, 6 modules are detected as relevant in the response to salt
stress in rice: 3 modules of 3 genes, each associated with shoot
$K$ content; 2 modules of 3 genes associated with shoot biomass; and 1
module of 4 genes associated with root biomass.

\subsection{Gene enrichment}

From the $19$ genes selected by LASSO, all but $3$ genes (the ones associated to $K$ content), are also identified as deferentially expressed ($|\ell_{ij}| \geq 2$) for at least one of the $92$ accessions. This suggest that those genes are strong candidates as stress responsive genes to salt conditions in rice. According to the Quickgo database, only $2$ of the $16$ diferentially expressed genes (both from the module related with shoot biomass) are named and have an associated protein product: Spermidine hydroxycinnamoyltransferase 2 (SHT2) and Lipoxygenase. Figure~\ref{fig:3d} shows their corresponding 3D protein structures.

% Figure 3d protein strucures
\begin{figure}[hptb]
  \centering
  \subfloat[Spermidine hydroxycinnamoyltransferase 2]{%
    \includegraphics[width=0.45\textwidth]{figures/structure_LOC_Os12g27254.png}
    \label{fig:structure_SHT2}
  }\hfill
  \subfloat[Lipoxygenase]{%
    \includegraphics[width=0.45\textwidth]{figures/structure_LOC_Os12g37260.png}
    \label{fig:structure_Lipoxygenase}
  }
  \caption{3D protein structure of named genes selected by LASSO, available from~\cite{szklarczyk2016string}.}
  \label{fig:3d}
\end{figure}

Uniprot database~\cite{uniprot2018uniprot} reports, on the one hand,
that SHT2 contributes to the natural variation of spermidine-based
phenolamides in rice cultivars. On the other hand, it is reported that
plant lipoxygenase may be involved in a number of diverse aspects of
plant physiology including growth and development, pest resistance,
and senescence or responses to wounding~\cite{uniprot2018uniprot}. This protein is involved in
the pathway oxylipin biosynthesis, which is part of Lipid
metabolism. Aditionally, previous studies
%
in~\cite{gupta2013plant,hou2015persimmon,mittova2002salt,peng2019novel,roychoudhury2011amelioration}
%
provide evidence of biological implications of sperimidine and
lipoxygenase in tolerance to salt stress in other plants or even in
rice cultivars. However, further studies are needed to elucidate the
detailed biological function of the remaining 14 genes that have not
been named so far, which may have a potential relevance in stress
responsive mechanisms to salt conditions in rice.

>>>>

%\section*{Results}
\section*{Identifying Potential Saline Stress Responsive Genes in Rice}
\label{sec.case}

This section presents a case study, applying the approach introduced in 
\hyperref[sec.framework]{the Workflow} section, to identify genes in
\textit{Oryza sativa} that respond to saline stress.
\vspace{0.5cm}

The RNA-seq data are available from the
GEO database \cite{clough2016gene} (accession
number GSE98455). It corresponds to $n_0=57\,845$ gene
expression profiles of shoot tissues measured for control and
salt conditions in $m=92$ accessions of the Rice Diversity Panel 1~\cite{eizenga2014registration},
with $r=2$ biological replicates. A total of $p=3 $ phenotypic traits
are used: shoot $K$ content, root biomass, and shoot biomass. These
traits were measured for the same $92$ genotypes, under control and
treatment conditions, and can be found in the supplementary
information for~\cite{campbell2017allelic}.

\subsection*{A. Data Pre-processing}

DESeq2 normalization is applied to the raw data and the biological
replicates are averaged. Genes exhibiting low variance are
identified as those with ratio of upper quantile to lower quantile
smaller than $1.5$ and are removed from the normalized data. Genes
with low expression, corresponding to those having more than $80\%$
samples with values smaller than $10$, are also removed. After this
filtering process a total of $n_1 = 9\,414$ genes are kept for further analysis.
\vspace{0.5cm}

From the Log Fold Change matrix $L_0$, genes whose difference between
upper quantile and lower quantile is greater than $0.25$ are
removed. Therefore, the resulting matrix $L_1$ contains the log ratios
of $n_2 = 8\,928$ genes. The logarithmic ratios of the phenotypic data,
for the $92$ accessions and the $3$ traits, are also computed.

\subsection*{B. Construction of the Co-expression Network}

The Log Fold Change matrix $L_1$ is used to compute the corresponding
similarity matrix.  For this network, it is observed that $\beta=3$ is
the smallest integer such that the $R^2 \geq 0.8
$. Figure~\ref{fig:beta} depicts the degree distribution of the
similarity matrix (left) and the degree distribution of the adjacency
matrix (right), which is the degree distribution of a scale-free
network with $R^2 = 0.8$ with $\beta = 3$.
\vspace{0.5cm}

%figure 2
\begin{figure}[htbp]
  \centering
    \includegraphics[clip,width=0.8\textwidth]{figures/figure2.png}
  \caption{Degree distribution of the network represented by $S$ (left) and $A$ (right).}
  \label{fig:beta}
\end{figure}

The resulting adjacency matrix $A$ represents a complete graph
$G=(V,E)$, with $|V| = 8\,928$ genes ($|E| = 39\,850\,128$ edges).

\subsection*{C. Identification of Co-expression Modules}

The adjacency matrix $A$ is transformed into an unweighted network
$\hat{A}$ applying the approach described
in~\cite{aoki2007approaches}. The cutoff value is set to $0.2$, 
based on the density of the network
combined with the decreasing number of nodes and edges with higher PCC
values. Thus keeping only the
connections above this threshold and removing any isolated nodes. The
resulting adjacency matrix $\hat{A}$ has $5\,810$ connected genes and
accounts for $614\,501$ edges.
\vspace{0.5cm}

After applying the HLC algorithm, a total of $4\,131$ genes are
distributed in $c = 5\,143$ overlapping modules of at least $3$
genes. Figure~\ref{fig:overlap} presents a histogram of the
overlapping percentage of these genes, measured as the proportion of
modules to which each gene belongs. The first bar of the histogram
represents the genes with zero overlap, corresponding to $28\%$ of the
total genes; the remaining $72\%$ represents the genes belonging to
more than one module.

%figure 3
\begin{figure}[htbp]
  \centering
    \includegraphics[clip,width=0.95\textwidth]{figures/figure3.pdf}
  \caption{Overlapping percentage of genes after applying HLC.}
  \label{fig:overlap}
\end{figure}

\subsection*{D. Detection of Modules Association to Phenotypic Traits}

The phenotypic traits under study are shoot $K$ content, root
biomass, and shoot biomass. Figure~\ref{fig:pdata} suggests that there
are significant differences in the values of these phenotypic traits
between stress and control conditions. This supports the working
hypothesis that these three variables represent tolerance-associated
traits in rice under salt stress.
\vspace{0.5cm}

%figure 4
\begin{figure}[htbp]
  \centering
    \includegraphics[clip,width=0.9\textwidth]{figures/figure4.pdf}
  \caption{Phenotipic traits distribution under control and salt stress.}
  \label{fig:pdata}
\end{figure}

Using the affiliation matrix $F$ derived from the HLC output and the
Log Fold Change matrix $L_1$, a matrix $M$ is built by computing the
eigengene for each of the $c = 5\,143$ modules. The LASSO technique is
applied by using each of the phenotypic traits as the outcome
variable, one at a time. As shown in Figure~\ref{fig:cross-val},
cross-validation is performed for each phenotypical trait in order to
select the corresponding regularization parameter $\lambda$ that
minimizes the mean-squared error.
\vspace{0.5cm}

%figure 5
\begin{figure}[htbp]
  \centering
    \includegraphics[clip,width=0.96\textwidth]{figures/figure5.pdf}
  \caption{Cross-validation of the LASSO regularization parameter
    $\lambda$, for each phenotypic trait.}
  \label{fig:cross-val}
\end{figure}

Finally, three LASSO models are adjusted by using the corresponding
$\lambda$ and phenotypical data with the eigengenes of matrix $M$. As
result, 6 modules are detected as relevant in the response to salt
stress in rice: 3 modules of 3 genes, each associated with shoot $K$
content; 2 modules of 3 genes associated with shoot biomass; and 1
module of 4 genes associated with root biomass (see
Figure~\ref{fig:final_genes}).

\subsection*{E. Gene Enrichment}

From the $19$ genes selected by LASSO, all but $3$ genes (the ones
associated to $K$ content), are also identified as deferentially
expressed ($|\ell_{ij}| \geq 2$) for at least one of the $92$
accessions. These genes are strong 
candidates target genes in rice.
%candidates as stress responsive genes to salt conditions in rice.
\vspace{0.5cm}

Figure~\ref{fig:final_genes} summarizes how from the initial
$n_0=57\,845$ genes, the proposed workflow identifies a reduced set of
$19$ genes. First, $48\,431$ genes are discarded after filtering the
normalized expression data $D_2$ and then $486$ additional genes are
discarded when filtering the Log Fold Change matrix $L_0$, to finally
arrive at $19$ genes, of which $16$ are differentially expressed.
\vspace{0.5cm}

% figure 6
\begin{figure}[htbp]
  \centering
    \includegraphics[clip,width=0.8\textwidth]{figures/figure6.pdf}
   \caption[Venn diagram for the case study in rice]%
   {Venn diagram representing the number of genes selected at
    different stages of the proposed workflow for the case study in
    rice.}
  \label{fig:final_genes}
\end{figure}

According to the Quickgo database, only $2$ of the $16$ differentially
expressed genes (both from the module related to shoot biomass) are
named and have an associated protein product: spermidine
hydroxycinnamoyltransferase 2 (SHT2) and lipoxygenase.
Figure~\ref{fig:3d} shows their corresponding 3D protein structures.
\vspace{0.5cm}

%figure 7
\begin{figure}[htbp]
  \centering
    \includegraphics[clip,width=0.8\textwidth]{figures/figure7.pdf}
  \caption{3D protein structure of named genes selected by LASSO, borrowed from~\cite{szklarczyk2016string}.}
  \label{fig:3d}
\end{figure}

The Uniprot database~\cite{uniprot2018uniprot} reports, on the one
hand, that SHT2 contributes to the natural variation of
spermidine-based phenolamides in rice cultivars. On the other hand, it
is reported in~\cite{uniprot2018uniprot} that plant lipoxygenase may
be involved in a number of diverse aspects of plant physiology
including growth and development, pest resistance, and senescence or
responses to wounding. This protein is involved in the pathway
oxylipin biosynthesis, which is part of Lipid metabolism. Additionally,
previous studies
%
in~\cite{gupta2013plant,hou2015persimmon,mittova2002salt,peng2019novel,roychoudhury2011amelioration}
%
provide evidence of biological implications of sperimidine and
lipoxygenase in tolerance to salt stress in other plants or even in
rice cultivars.

As a conclusion, the results presented in this section suggest that
further studies are needed to elucidate the detailed biological
function of the remaining 14 genes that have not been named so far in
the literature.  They may have the potential to intervene in stress
responsive mechanisms to salt conditions in rice.

### 2020-12-07 camila/report/concl.tex
references added
bmc version

\section{Concluding Remarks}
\label{sec.concl}

This manuscript provides a detailed description of a network-based analysis
methodology for the discovery of key genes responding to
a specific treatment in an organism. It links transcriptomic with
phenotypic data, and identifies overlaping gene modules.

The proposed methodology is inspired by the workflow suggested in the
WGCNA methodology~\cite{langfelder2008wgcna}. The main steps are the
preprocessing of the gene expression data, the construction of a
co-expression network, the detection of modules within the network,
the relation of modules with external information (e.g. phenotypic data)
and the enrichment of the identified key genes with additional information.
The WGCNA methodology and therefore the proposal as well, are structured
in a modular way, which allows modifying and  exploring different techniques
in each step of the process.

The proposed approach is designed to integrate expression data
measured under two different conditions (namely, control and treatment),
unlike the usually co-expression-based approaches which work with both
conditions independently or consider only a single condition. For this
purpose, an approach similar to that proposed in~\cite{du2019network}
is used, where the control and treatment data are compiled in a single
matrix using the log fold change measure. Thus, the input to construct
the co-expression network is not the expression data, but instead the changes
in the expression levels from one condition to the other, making room for capturing
the signal of changes caused by the treatment.

Another important feature in the proposed methodology is the module
detection technique. The co-expression network is computed as in
WGCNA until a scale-free network is obtained. This network is then
used to apply the HLC algorithm, a clustering technique capable of
detecting overlapping communities. Several approaches of module
detection from gene expression have been proposed and were evaluated
in~\cite{saelens2018comprehensive}. Most of them focuses only on
disjoint (non-overlapping) communities, and the described techniques
dealing with overlaps are not clustering but biclustering and
decomposition methods. It is well known that communities in real
networks are overlapping~\cite{palla2005uncovering}. Thus, the approach
presented in this work can be seen as a generalization of the previous 
approaches such as WGCNA.

The proposed methodology was applied in a case study with rice under salt
stress. The results show a group of 14 genes in which $2$ of them are
related to the response to saline stress according to previous
studies, validating the ability of the method to detect this kind of
key genes. As future work, other overlapping module detection and selection
techniques should be used instead HLC and LASSO, respectively. The
combination of these techniques would allow finding target genes for
future biological studies that evaluate their potential as genes that
respond to salt stress in rice. Also, laboratory experimentation needs to be
conducted on via to verify the findings of this paper in relation to salinity 
stress. Finally, this study can be extended to other stresses and even to other 
crops.

>>>>

\section*{Concluding Remarks}
\label{sec.concl}

This manuscript provides a detailed description of a network-based
analysis workflow for the discovery of key genes responding to a
specific treatment in an organism. It links transcriptomic with
phenotypic data and identifies overlapping gene modules.
\vspace{0.5cm}

The proposed approach is inspired by the workflow suggested in the
WGCNA~\cite{langfelder2008wgcna}. Its main steps are the preprocessing
of the gene expression data, the construction of a co-expression
network, the detection of modules within the network, the relation of
modules with external information (e.g., phenotypic data), and the
enrichment of the identified key genes with additional information.
Both approaches are structured in a modular way, which allows
modifying and exploring different techniques in each step of the
workflow.
\vspace{0.5cm}

The proposed workflow is designed to integrate expression data
measured under two different conditions (namely, control and
treatment), unlike the usually co-expression-based approaches which
work with both conditions independently or consider only a single
condition. For this purpose, an approach similar to that proposed
in~\cite{du2019network} is used, where the control and treatment data
are compiled in a single matrix using the Log Fold Change
measure. Thus, the input to construct the co-expression network is not
the expression data, but instead the changes in the expression levels
from one condition to the other, making room for capturing the signal
of changes caused by the treatment.
\vspace{0.5cm}

An important feature in the proposed workflow is the module
detection technique. The co-expression network is computed, as in
WGCNA, until a scale-free network is obtained. In the proposed
approach, this network is then used to apply the HLC algorithm, a
clustering technique capable of detecting overlapping
communities. Several approaches of module detection from gene
expression have been proposed and were evaluated
in~\cite{saelens2018comprehensive}. Most of them focus mainly on
disjoint (non-overlapping) communities; the techniques described
dealing with overlaps are not clustering, but bi-clustering and
decomposition methods. It is well known that communities in real
networks, including biological ones, are likely
overlap~\cite{palla2005uncovering}. Thus, the approach presented
in this work can be seen as a generalization of the previous
approaches, such as WGCNA, with the potential to deal with genes
associated to multiple biological processes.
\vspace{0.5cm}

The approach was applied in a case study with rice under salt
stress. The results show a group of 14 genes, of which only $2$ of
them have been previously related to saline stress response in other
studies. As future work, other overlapping module detection and
selection techniques should be used instead HLC and LASSO,
respectively. The combination of these techniques would allow finding
target genes for future biological studies that evaluate their
potential as genes that respond to salt stress in rice, and other
crops and stresses. In-vivo laboratory experimentation needs to be
conducted to validate the findings of this paper in relation to
salinity stress.
\vspace{0.5cm}

Finally, the workflow is presented as a protocol capable of
considerably reducing the number of genes detected as relevant in the
response to stress of choice. Other traditionally used methods for
this purpose tend to generate a large list of candidate genes, thus
limiting subsequent efforts in experimental validation. In this sense,
the proposed workflow can help in reducing such efforts in time and
money invested by researchers in the experimental validation of
stress-responsive genes.
